Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes

### 2017-05-06 pasp/paperclimate.tex
oscar and my corrections

\title[Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing]{ Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes}
% Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes

\author{Germ¬\'an Chaparro Molano$^{1}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\ead{gchaparrom@ecci.edu.co}
\author{Oscar Leonardo Ram\'irez Su\'arez$^{1}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\ead{oramirezs@ecci.edu.co}
\author{Oscar Restrepo$^{1}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\ead{orestrepog@ecci.edu.co}
\author{Alexander Mart\'inez$^{1,2}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\address{$^2$Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales, Bogot\'a, Colombia}
\vspace{10pt}
\begin{indented}
\item[]December 2016
\end{indented}

@@

The development of astronomical instrumentation technology in the 0.2-2 THz range has been rapidly growing in recent years. Also, interest in THz technology, specially applied to satellite telecommunications has also grown due to the saturation of the ``classical radio window'' in telecommunications  \cite{newera}, and atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal} have been developed recently due to this interest. Near-equatorial countries face a challenge in using this frequency band for astronomical observations and telecommunications using terabit satellite links  \cite{suen2016} due to the presence of a tropical belt of dense water vapor which efficiently absorbs THz radiation \cite{tropicalbelt}. In the Northern Andes, previous studies of astronomical site testing in the visible range have shown that high nighttime humidity conditions make this region suitable only for educational observatories \cite{pinzon}. However, considering that millimeter/sub-millimeter observations need not be done during the night, unusually dry, clear-daytime-sky, high altitude regions in the Northern Andes could be suitable candidates for a world-class millimeter-wave observatory. \\

@@

The water vapor column above a potential site can change rapidly (during the course of the day) or it can change seasonally \cite{arm2013,caumont2016}. This is why medium-to-long term monitoring of the atmosphere above a site is required before building an expensive mm/sub-mm wave radio telescope. However, even in best-case scenarios, earth-based telescopes will always be limited by atmospheric absorption of light specially at THz frequencies, where windows of observations are few and in some cases, narrow \cite{arrlim,limits}. \\

Although radiometer measurements are desirable for directly measuring the atmospheric water vapor column, their development and/or deployment can be complex and expensive \cite{radiopro,receiver}. For this reason we screened historic climate data for evidence of local long-term clear sky, low humidity conditions as proxies for a locally low precipitable water vapor column. By low-dimensionally embedding the data, we are able to reduce the dimensionality of multi-annual monthly data from $N_\mathrm{dim}=12$ to $N_\mathrm{dim}\le3$. Thus, we can correlate unusually dry regions regardless of their geographical location. Given that the climatological variables were sparsely measured by weather stations across Colombia, we evaluated the aptness of a given location using a Bayesian probability-derived quality index. Finally, we geographically clustered our candidate locations in order to identify regions of interest. We compared our regions of interest with regions with unusually low precipitable water vapor in low-resolution water vapor satellite maps \cite{suen2016} and find a possibly significative geographical correlation.  In an upcoming paper we will analize satellite data in order to retrieve high-resolution seasonal precipitable water vapor maps for our regions of interest.\\

@@

The meteorological data used in this study were gathered over 30 years (1981-2010) in 2046 weather stations monitored by the Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales (IDEAM) in Colombia (Figure \ref{totmap}). The  variables that we considered relevant to our work were: Precipitation (mm/mo), Rain Days (d/mo),  Relative Humidity (\%), and Sunshine Duration (h/d). These variables are reported as multi-annual monthly data, i.e. each variable is reported monthly from January to December averaged over the 30 year range. For precipitation, each monthly datum corresponds to the cumulative monthly value.  For rain days, each monthly datum corresponds to the number of rain days for that month.  For relative humidity, each monthly datum corresponds to the  average daily value averaged over each month. For sunshine duration, each monthly datum corresponds to the daily value averaged over each month. All stations reported precipitation values, although only 2002 reported rain days, 445 reported relative humidity, and 336 reported sunshine duration. This means that not all stations registered all the meteorological variables relevant to this work. Thus, we classify stations according to which variables they measured (Table \ref{tabstype}).\\

@@

\begin{table}
\caption{\label{tabstype}Types of stations classified according to its measured variables.}
\begin{indented}
\item[]\begin{tabular}{@{}cccccc}
\br
Type&Precip.&Rain &Relative&Sunshine&No. of Stations\\
&&Days&Humidity&Duration&\\
\mr
$T_1$&Y&N&N&N&29\\
$T_2$&Y&Y&N&N&1563\\
$T_3$&Y&N&Y&N&1\\
$T_4$&Y&Y&Y&N&117\\
$T_5$&Y&N&N&Y&1\\
$T_6$&Y&Y&N&Y&8\\
$T_7$&Y&N&Y&Y&13\\
$T_8$&Y&Y&Y&Y&314\\
\br
\end{tabular}
\end{indented}
\end{table}

@@

We classified stations using a low-dimensional embedding of the data \cite{embed} for each variable. This ensures that a classification algorithm can uncover features in the data even even when the data are projected across a number of dimensions that is lower than the native dimensionality of the data (which in this case is monthly, i.e. $N_\mathrm{dim}=12$). This involves applying a dimensionality reduction algorithm followed by a clustering algorithm. Given that we wish to make a grouping of stations showing similar climatological patterns without making too many assumptions on the data, we decided to use unsupervised learning techniques.  In our case, we chose Principal Component Analysis (PCA) covering 95\% of the variance followed by a Gaussian Mixture Model selected by a low/locally minimum Bayesian Information Criterion (BIC).

@@

We were able to reduce the dimensionality of the data for each variable from  $N_\mathrm{dim}=12$ to  $N_\mathrm{dim}^\mathrm{PCA}\le3$, while ensuring that at least 95\% of the variance of the data is explained by the least amount of components. For precipitation, rain days and sunshine duration, $N_\mathrm{dim}^\mathrm{PCA}(95\%)=3$ and for relative humidity $N_\mathrm{dim}^\mathrm{PCA}(95\%)=2$. To do this we used the Python \texttt{sklearn.decomposition.PCA} module \cite{sklearn}. Figure \ref{pcah} shows the relative humidity data projected across 2 principal components explaining 95\% of the variance, thus preserving most of the original structure of the data.

@@

After projecting the data along the components found by the PCA algorithm, we clustered the dimensionally-reduced data using a Gaussian Mixture Model, which uses an Expectation-Maximization algorithm to find a Maximum Likelihood model composed of a number $N_\mathrm{G-C}$ of Gaussian distributions, each of which represents a cluster in the data \cite{gmm}. Depending on the restrictions put on the covariance of each Gaussian distribution across the data space, the number of free parameters can go from $N_\mathrm{G-C}(N_\mathrm{dim}+1)$ (spherical covariance) to  $N_\mathrm{G-C}(3N_\mathrm{dim}+1)$ (full covariance). Since the number of components and covariance restriction is given by the user and not \emph{a posteriori} by the algorithm, me need to make sure that the Gaussian components do not over-fit the data, e.g. a model which yields one cluster per datum should be disallowed.\\

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{bich.pdf}
\includegraphics[scale=0.5]{bicl.pdf}\\
\includegraphics[scale=0.5]{bicd.pdf}
\includegraphics[scale=0.5]{bicb.pdf}
\caption{Bayesian Information Criterion (BIC) results for Gaussian Mixture Models using a different number of clusters and using different covariance restriction methods applied on each measured variable dataset.}\label{bic}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{gmmh0.pdf}
\includegraphics[scale=0.5]{gmmh1.pdf}
\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for relative humidity. Left column: Average monthly relative humidity and standard deviation (error bars) for each climate cluster. Right column: Geographical location and altitude of weather stations in each relative humidity cluster.}\label{clusth}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{gmml0.pdf}
\includegraphics[scale=0.5]{gmml1.pdf}
\includegraphics[scale=0.5]{gmml2.pdf}
\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for precipitation. Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster. Right column: Geographical location and altitude of weather stations in each precipitation cluster.}\label{clustl}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{gmml3.pdf}
\includegraphics[scale=0.5]{gmml4.pdf}
\includegraphics[scale=0.5]{gmml5.pdf}
\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for precipitation. Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster. Right column: Geographical location and altitude of weather stations in each precipitation cluster.}
\end{center}
\end{figure}

@@

\includegraphics[scale=0.5]{gmml6.pdf}
\includegraphics[scale=0.5]{gmml7.pdf}
\includegraphics[scale=0.5]{gmml8.pdf}
\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for precipitation. Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster. Right column: Geographical location and altitude of weather stations in each precipitation cluster.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{gmml9.pdf}
\includegraphics[scale=0.5]{gmml10.pdf}
\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for precipitation. Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster. Right column: Geographical location and altitude of weather stations in each precipitation cluster.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{gmmd0.pdf}
\includegraphics[scale=0.5]{gmmd1.pdf}
\includegraphics[scale=0.5]{gmmd2.pdf}
\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for rain days. Left column: Average monthly rain days and standard deviation (error bars) for each climate cluster. Right column: Geographical location and altitude of weather stations in each rain days cluster.}\label{clustd}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{gmmd3.pdf}
\includegraphics[scale=0.5]{gmmd4.pdf}
\includegraphics[scale=0.5]{gmmd5.pdf}
\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for rain days. Left column: Average monthly rain days and standard deviation (error bars) for each climate cluster. Right column: Geographical location and altitude of weather stations in each rain days cluster.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{gmmb0.pdf}
\includegraphics[scale=0.5]{gmmb1.pdf}
\includegraphics[scale=0.5]{gmmb2.pdf}
\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for sunshine duration. Left column: Average monthly sunshine duration and standard deviation (error bars) for each climate cluster. Right column: Geographical location and altitude of weather stations in each sunshine duration cluster.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{gmmb3.pdf}
\includegraphics[scale=0.5]{gmmb4.pdf}
\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for sunshine duration. Left column: Average monthly sunshine duration and standard deviation (error bars) for each climate cluster. Right column: Geographical location and altitude of weather stations in each sunshine duration cluster.}\label{clustb}
\end{center}
\end{figure}

\begin{table}
\caption{\label{tabclu}Climate clusters according to a lowest BIC-based selection of Gaussian Mixture Models. See Figure \ref{bic}.}
\begin{indented}
\item[]\begin{tabular}{@{}lclll}
\br
Variable&Symbol&Lowest BIC&Lowest BIC&Preferred Clusters\\
&&No. of Clusters&Cov. Method &\\
\mr
Precipitation &$R$&11&Full&2,3,7,11\\
Rain Days&$D$&6&Full&1,4,6\\
Relative Humidity&$H$&2&Full&2\\
Sunshine Duration&$S$&5&Tied&1,4,5\\
\br
\end{tabular}
\end{indented}
\end{table}

@@

We wish to identify all stations for whom all four criteria (humidity, precipitation, rain days, sunshine duration) for identifying a region as having a stable sunny, dry climate can be met.  However, since only a few stations measured all variables, it is not straightforward to accept or reject stations based on their satisfing a criterion, i.e. belonging to one of our preferred clusters. This requires setting up a measure of quality for individual stations depending on their location and the variables they measured. We will describe the selection process and probabilistic analysis that allows us to quantify a station's capability for indicating regions with an unusually dry, sunny climate.\\

@@

We quantify the quality of a given site using the probability for the hypothesis that a station which appears in our shortlist meets all four criteria, given that it is located at a given elevation. We can write down this probability in terms of the following events:

\begin{itemize}
\item $C$ is the event that a given station meets all four criteria, i.e. belongs to our preferred clusters for the four variables relevant to this study.
\begin{itemize}
\item $C$ is actually the conjunction of all four of the $R,D,H,S$ events ``The station belongs to a preferred cluster for variable $X$'', i.e. $C=R\cap D\cap H \cap S$.
\end{itemize}
\item $A$ is the event that a given station appears in our shortlist.
\item $T_i$ is the event that a given station is of type $T_i$ (See Table \ref{tabstype}).
\begin{itemize}
\item For example, if $M_X$ is the event ``the station measured variable $X$'', then the event $T_7$ is equivalent to $M_R\cap \neg M_D\cap M_H \cap M_S$
\end{itemize}
\item $h$ is the event that a given station is located at an elevation $h$.
\end{itemize}

>>>>

\title[Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing]{Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes}
% Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes

\author{Germ¬\'an Chaparro Molano$^{1}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\ead{gchaparrom@ecci.edu.co}
\author{Oscar Leonardo Ram\'irez Su\'arez$^{1}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\ead{oramirezs@ecci.edu.co}
\author{Oscar Restrepo$^{1,2}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\address{$^2$Radio Astronomy Instrumentation Group, Universidad de Chile, Santiago de Chile, Chile}
\ead{orestrepog@ecci.edu.co}
\author{Alexander Mart\'inez$^{1,3}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\address{$^3$Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales, Bogot\'a, Colombia}
\vspace{10pt}
\begin{indented}
\item[]December 2016
\end{indented}

@@

The development of astronomical instrumentation technology in the 0.2-2 THz range has been rapidly growing in recent years. Another recent surge in interest in telecommunications THz technology has appeared recently due to the saturation of the ``classical radio window'' in telecommunications  \cite{newera}. For this reason, there has been development of atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal}. Near-equatorial countries face a challenge in using this frequency band for astronomical observations and telecommunications using terabit satellite links  \cite{suen2016} due to the presence of a tropical belt of dense water vapor which efficiently absorbs THz radiation \cite{tropicalbelt}. In the northern Andes, previous studies of astronomical site testing in the visible range have shown that high nighttime humidity conditions make this region suitable only for educational observatories \cite{pinzon}. However, considering that millimeter/sub-millimeter observations need not be done during the night, unusually dry, clear-daytime-sky, high altitude regions in the Northern Andes could be suitable candidates for a world-class millimeter-wave observatory. \\

@@

The water vapor column above a potential site can change rapidly (during the course of the day) or it can change seasonally \cite{arm2013,caumont2016}. This is why medium-to-long term monitoring of the atmosphere above a site is required before building an expensive mm/sub-mm wave radio telescope. However, even in best-case scenarios, earth-based telescopes will always be limited by atmospheric absorption of light specially at THz frequencies, where frequency windows of observations are few and in some cases, narrow \cite{arrlim,limits}. \\

Although radiometer measurements are desirable for directly measuring the atmospheric water vapor column, their development and/or deployment can be complex and expensive \cite{radiopro,receiver}. For this reason we screened historic climate data for evidence of local long-term clear sky, low humidity conditions as proxies for a locally low precipitable water vapor column. By low-dimensionally embedding the data, we are able to reduce the dimensionality of multi-annual monthly data from $N_\mathrm{dim}=12$ to $N_\mathrm{dim}\le3$ while preserving $>95$\% of the variance in the data. Thus, we can correlate unusually dry regions regardless of their geographical location. Given that the climatological variables were sparsely measured by weather stations across Colombia, we evaluated the aptness of a given location using a Bayesian probability-derived quality index. Finally, we geographically clustered our candidate locations in order to identify regions of interest. We compared these regions of interest with regions with unusually low precipitable water vapor in low-resolution ($\sim20$ km) water vapor satellite maps \cite{suen2016} and find a possibly significative geographical correlation.  In an upcoming paper we will analize satellite data in order to retrieve high-resolution seasonal precipitable water vapor maps for the regions of interest reported here.\\

@@

The meteorological data used in this study were gathered over 30 years (1981-2010) in 2046 weather stations monitored by the Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales (IDEAM) in Colombia (Figure \ref{totmap}). The  variables that we considered relevant to our work were: Elevation (in meters above sea level, or masl), Precipitation (mm/mo), Rain Days (d/mo),  Relative Humidity (\%), and Sunshine Duration (h/d). The climatological variables are reported as multi-annual monthly data, i.e. each variable is reported monthly from January to December averaged over the 30 year range. For precipitation, each monthly datum corresponds to the cumulative monthly value.  For rain days, each monthly datum corresponds to the number of rain days for that month.  For relative humidity, each monthly datum corresponds to the  average daily value averaged over each month. For sunshine duration, each monthly datum corresponds to the daily value averaged over each month. All stations reported precipitation values, although only 2002 reported rain days, 445 reported relative humidity, and 336 reported sunshine duration. This means that not all stations registered all the meteorological variables relevant to this work. Thus, we classify stations according to which variables they measured (Table \ref{tabstype}).\\

@@

\begin{table}
\caption{\label{tabstype}Types of stations classified according to its measured variables. Y = measured, N = not measured.}
\begin{indented}
\item[]\begin{tabular}{@{}cccccc}
\br
Type&Precip.&Rain &Relative&Sunshine&No. of Stations\\
&&Days&Humidity&Duration&\\
\mr
$T_1$&Y&N&N&N&29\\
$T_2$&Y&Y&N&N&1563\\
$T_3$&Y&N&Y&N&1\\
$T_4$&Y&Y&Y&N&117\\
$T_5$&Y&N&N&Y&1\\
$T_6$&Y&Y&N&Y&8\\
$T_7$&Y&N&Y&Y&13\\
$T_8$&Y&Y&Y&Y&314\\
\br
\end{tabular}
\end{indented}
\end{table}

@@

We classified stations using a low-dimensional embedding of the data \cite{embed} for each variable. This ensures that a classification algorithm can uncover features in the data even when the data are projected across a number of dimensions that is lower than the native dimensionality of the data (which in this case is monthly, i.e. $N_\mathrm{dim}=12$). This involves applying a dimensionality reduction algorithm followed by a clustering algorithm. Given that we wish to make a grouping of stations showing similar climatological patterns without making assumptions on the data, we decided to use unsupervised learning techniques.  In our case, we chose Principal Component Analysis (PCA) covering 95\% of the variance followed by a Gaussian Mixture Model selected by a low/locally minimum Bayesian Information Criterion (BIC).

@@

We were able to reduce the dimensionality of the data for each variable from  $N_\mathrm{dim}=12$ to  $N_\mathrm{dim}^\mathrm{PCA}\le3$, while ensuring that at least 95\% of the variance of the data is explained by the least amount of components. For precipitation, rain days and sunshine duration, $N_\mathrm{dim}^\mathrm{PCA}(95\%)=3$ and for relative humidity $N_\mathrm{dim}^\mathrm{PCA}(95\%)=2$. To do this we used the Python \texttt{sklearn.decomposition.PCA} module \cite{sklearn}. To illustrate, Figure \ref{pcah} shows the relative humidity data projected across 2 principal components explaining 95\% of the variance, thus preserving most of the original structure of the data. Dimensionality-reduced precipitation, rain days, and sunshine duration data are shown later (Figure \ref{lde}).

@@

After projecting the data along the components found by the PCA algorithm, we clustered the dimensionally-reduced data using a Gaussian Mixture Model, which uses an Expectation-Maximization algorithm to find a Maximum Likelihood model composed of a number $N_\mathrm{G-C}$ of Gaussian distributions, each of which represents a cluster in the data \cite{gmm}. Depending on the restrictions put on the covariance of each Gaussian distribution across the data space, the number of free parameters can go from $N_\mathrm{G-C}(N_\mathrm{dim}+1)$ (spherical covariance) to  $N_\mathrm{G-C}(3N_\mathrm{dim}+1)$ (full covariance). Since the number of components and covariance restriction is given by the user and not \emph{a posteriori} by the algorithm, we need to make sure that the Gaussian components do not over-fit the data, e.g. a model which yields one cluster per datum should be disallowed.\\

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{bich.pdf}
\includegraphics[scale=0.5]{bicl.pdf}\\
\includegraphics[scale=0.5]{bicd.pdf}
\includegraphics[scale=0.5]{bicb.pdf}
\caption{Bayesian Information Criterion (BIC) results for Gaussian Mixture Models using a different number of clusters and using different covariance restriction methods \cite{sklearn} applied on each measured variable dataset.}\label{bic}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{gmmh0.pdf}
\includegraphics[scale=0.5]{gmmh1.pdf}
\caption{Climate clusters (1-2 of 2) as predicted by the lowest-BIC Gaussian Mixture Model for relative humidity. Left column: Average monthly relative humidity and standard deviation (error bars) for each climate cluster. Right column: Geographical location and altitude of weather stations in each relative humidity cluster.}\label{clusth}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{gmml0.pdf}
\includegraphics[scale=0.5]{gmml1.pdf}
\includegraphics[scale=0.5]{gmml2.pdf}
\caption{Climate clusters (1-3 of 11) as predicted by the lowest-BIC Gaussian Mixture Model for precipitation. Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster. Right column: Geographical location and altitude of weather stations in each precipitation cluster.}\label{clustl}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{gmml3.pdf}
\includegraphics[scale=0.5]{gmml4.pdf}
\includegraphics[scale=0.5]{gmml5.pdf}
\caption{Climate clusters  (4-6 of 11) as predicted by the lowest-BIC Gaussian Mixture Model for precipitation. Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster. Right column: Geographical location and altitude of weather stations in each precipitation cluster. Continuation of Figure \ref{clustl}.}\label{clustl2}
\end{center}
\end{figure}

@@

\includegraphics[scale=0.5]{gmml6.pdf}
\includegraphics[scale=0.5]{gmml7.pdf}
\includegraphics[scale=0.5]{gmml8.pdf}
\caption{Climate clusters (7-9 of 11) as predicted by the lowest-BIC Gaussian Mixture Model for precipitation. Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster. Right column: Geographical location and altitude of weather stations in each precipitation cluster. Continuation of Figure \ref{clustl2}.}\label{clustl3}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{gmml9.pdf}
\includegraphics[scale=0.5]{gmml10.pdf}
\caption{Climate clusters (10-11 of 11) as predicted by the lowest-BIC Gaussian Mixture Model for precipitation. Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster. Right column: Geographical location and altitude of weather stations in each precipitation cluster. Continuation of Figure \ref{clustl3}.}\label{clustl4}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{gmmd0.pdf}
\includegraphics[scale=0.5]{gmmd1.pdf}
\includegraphics[scale=0.5]{gmmd2.pdf}
\caption{Climate clusters (1-3 of 6) as predicted by the lowest-BIC Gaussian Mixture Model for rain days. Left column: Average monthly rain days and standard deviation (error bars) for each climate cluster. Right column: Geographical location and altitude of weather stations in each rain days cluster.}\label{clustd}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{gmmd3.pdf}
\includegraphics[scale=0.5]{gmmd4.pdf}
\includegraphics[scale=0.5]{gmmd5.pdf}
\caption{Climate clusters (4-6 of 6) as predicted by the lowest-BIC Gaussian Mixture Model for rain days. Left column: Average monthly rain days and standard deviation (error bars) for each climate cluster. Right column: Geographical location and altitude of weather stations in each rain days cluster. Continuation of Figure \ref{clustd}.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{gmmb0.pdf}
\includegraphics[scale=0.5]{gmmb1.pdf}
\includegraphics[scale=0.5]{gmmb2.pdf}
\caption{Climate clusters (1-3 of 5) as predicted by the lowest-BIC Gaussian Mixture Model for sunshine duration. Left column: Average monthly sunshine duration and standard deviation (error bars) for each climate cluster. Right column: Geographical location and altitude of weather stations in each sunshine duration cluster.}\label{clustb1}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{gmmb3.pdf}
\includegraphics[scale=0.5]{gmmb4.pdf}
\caption{Climate clusters (4-5 of 5) as predicted by the lowest-BIC Gaussian Mixture Model for sunshine duration. Left column: Average monthly sunshine duration and standard deviation (error bars) for each climate cluster. Right column: Geographical location and altitude of weather stations in each sunshine duration cluster. Continuation of Figure \ref{clustb1}.}\label{clustb}
\end{center}
\end{figure}

\begin{table}
\caption{\label{tabclu}Climate clusters according to a lowest BIC-based selection of Gaussian Mixture Models following Figure \ref{bic}. The selected clusters column refers to the climate clusters shown in Figures \ref{clusth} to \ref{clustd}. }
\begin{indented}
\item[]\begin{tabular}{@{}lclll}
\br
Variable&Symbol&Lowest BIC&Lowest BIC&Selected Clusters\\
&&No. of Clusters&Cov. Method &\\
\mr
Relative Humidity&$H$&2&Full&2\\
Precipitation &$R$&11&Full&2,3,7,11\\
Rain Days&$D$&6&Full&1,4,6\\
Sunshine Duration&$S$&5&Tied&1,4,5\\
\br
\end{tabular}
\end{indented}
\end{table}

@@

We wish to identify all stations for whom all four criteria (humidity, precipitation, rain days, sunshine duration) for identifying a region as having a stable sunny, dry climate can be met.  However, since only a few stations measured all variables, it is not straightforward to accept or reject stations based on their satisfying a criterion, i.e. belonging to one of our preferred clusters. This requires setting up a measure of quality for individual stations depending on their location and the variables they measured. We will describe the selection process and probabilistic analysis that allows us to quantify a station's capability for indicating regions with an unusually dry, sunny climate.\\

@@

We quantify the quality of a given site using the probability for the hypothesis that a station $j$ which appears in our shortlist meets all four criteria, given that it is located at a given elevation. We can write down this probability $P_j$ in terms of the following events:

\begin{itemize}
\item $C$ is the event that a given station meets all four criteria, i.e. belongs to our preferred clusters for the four variables relevant to this study.
\begin{itemize}
\item $C$ is actually the conjunction of all four of the $R,D,H,S$ events ``the station belongs to a selected cluster for variable $X$'', i.e. $C=R\cap D\cap H \cap S$ (See Table \ref{tabclu}).
\end{itemize}
\item $A$ is the event that a given station appears in our shortlist.
\item $T_i$ is the event that a given station is of type $T_i$ (See Table \ref{tabstype}).
\begin{itemize}
\item For example, if $M_X$ is the event ``the station measured variable $X$'', then the event $T_7$ is equivalent to $M_R\cap \neg M_D\cap M_H \cap M_S$
\end{itemize}
\item $h$ is the event that a given station is located at an elevation $h$.
\end{itemize}

Thus, the probability that a station $j=(0,1,...,N_\mathrm{stations})$ of type $T_i$, meets all four criteria ($C$) and appears in our shortlist ($A$) given that it is located at an elevation $h$, can be written as $P_j=P(C\cap A\cap T_i\ |\ h)$. We cannot  calculate exactly this probability, as it depends on unknown unknowns, i.e. whether a station that did not measure a given variable in reality belongs to one of our preferred climatological clusters for that variable (Table \ref{tabclu}). Thus we will use Bayes' theorem to help estimate this probability using simpler conditional probabilities. First we control for the inhomogeneous distribution of elevations $P(h)$,
\begin{equation}
P_j=P(C\cap A\cap T_i\ |\ h)=\frac{P(C\cap A\cap T_i\cap h)}{P(h)}\ .
\end{equation}
The joint probability of all events $C\cap A\cap T_i\cap h$ can be rewritten as,
\begin{equation}
P_j=\frac{P(h\ |\ C\cap A\cap T_i)}{P(h)}P(C\cap A\cap T_i)\ .
\end{equation}
If a station satisfies all four criteria, $P(C\cap A\cap T_i)=1$. This probability can be expressed in terms of simpler conditional probabilities,
\begin{equation}\label{peejay}
P_j=\frac{P(h\ |\ C\cap A\cap T_i)}{P(h)}P(C\ |\ A\cap T_i)P(A\ |\ T_i)P(T_i)\ ,
\end{equation}
where 
\begin{itemize}
\item $P(h\ |\ C\cap A\cap T_i)$ is the distribution of elevations for a station type ($T_i$) on our shortlist ($C\cap A$), 
\item $P(A\ |\ T_i)$ is the probability that a station of type $T_i$ is in our shortlist, 
\item $P(T_i)$ is the probability that a station type is $T_i$, and 
\item $P(C\ |\ A\cap T_i)$ is the probability that a station meets all four criteria ($C$) given that it is in our shortlist ($A$) and is of type $T_i$. 
\end{itemize}
All of these probabilities except for the last one can be computed directly from the data\footnote{Distributions of elevation-related probabilities were approximated discretely using a bin size estimated from the mean elevation difference between stations when sorted by elevation. }.\\

 The probability $P(C\ |\ A\cap T_i)$ has to be estimated indirectly, as we cannot presume to know under which conditions a station can satisfy all our criteria. However, we can rewrite this probability as,
\begin{eqnarray}
P(C\ |\ A\cap T_i)&=P(R\cap D\cap H\cap S\ |\ A\cap T_i)\ , \\
\label{pclanti}
&=\frac{P(R\cap D\cap H\cap S\cap A\cap T_i)}{P(A\cap T_i)}\ . 
\end{eqnarray}
In order to estimate $P(C\ |\ A\cap T_i)$ we assume that the probability of a station meeting our criteria for a given number of variables is independent of the station type. To illustrate this, let us assume that a given station is of type 1, so it only measured precipitation and rain days, i.e.,
\begin{equation}
T_1=M_R\cap M_D\cap\neg M_H\cap\neg M_S\ .
\end{equation}
If this station of type $T_1$ is in our shortlist ($A$), the $R$ and $D$ criteria are met. Thus,
\begin{equation}
P(R\cap D\cap H\cap S\cap A\cap T_1)=P( H\cap S\ |\ A\cap T_1)\ .
\end{equation}
Substituting this into Eq. (\ref{pclanti}),
\begin{equation}
P(C\ |\ A\cap T_1)=\frac{P( H\cap S\ |\ A\cap T_1)}{P(A\cap T_1)}=P(H\cap S\ |\ A\cap T_1) \ .
\end{equation}
The last term in the previous expression is unknown, but we can approximate it assuming, as mentioned above, that it is independent of the station type. Therefore, 
\begin{equation}
P(C\ |\ A\cap T_)\simeq P(H\cap S)\ .
\end{equation}
The probability on the right hand side of the previous equation can be directly computed from the data, as it only requires counting how many stations meet both $H$ and $S$ criteria. This procedure can be extended to all other station types.\\

The probabilities for stations in our shortlist ($P_j$ in Equation \ref{peejay}) can be used to evaluate the quality of a site, but since they vary by orders of magnitude, for the sake of clarity we decided to use instead a probability-derived logarithmic quality index. Thus, we define the quality index $Q_j$ for a station $j$ in our shortlist as,
\begin{equation}
Q_j=9\ \frac{\log (\nicefrac{P_\mathrm{j}}{P_\mathrm{min}})}{\log (\nicefrac{P_\mathrm{max}}{P_\mathrm{min}})}+1\ .
\end{equation}
Here $P_\mathrm{min},P_\mathrm{max}$ are the lowest and highest values for the probabilities for the stations in the first shorlist of 665 stations that satisfy criteria for all measured variables. Thus, if $P_j=P_\mathrm{max}$, $Q_j=10$, and if $P_j=P_\mathrm{min}$, $Q_j=1$.
